---
title: "Final project2"
author: "Xiaoyan Wen"
date: "12/11/2020"
output: word_document
---
# Dataset 2:
# a) Mnemiopsis_col_data.csv
# b) Mnemiopsis_count_data.csv``

```{r}
rm(list = ls())
graphics.off()
```

```{r}
library(readr)
library(dplyr)
library(tibble)
library(ggbiplot) # for pca plotting
library(ggplot2)
library(pls)   # for pcr analysis
library(abind)
library(nnet) # for multinom regression
library(MASS) # for polr regression, stepwise regression
library(DAAG) # for cross-validation
library(dendextend)
library(factoextra)
library(cluster) # different methods of clustering
library(purrr) # for map function
library(caret) # split training and test set
library(neuralnet)
library(nnet)
library(treeio)
library(ggtree)
library(pheatmap)
library(DESeq2)



# load data
Mnemiopsis_count_data <- read_csv("Mnemiopsis_count_data.csv")
mncount <- column_to_rownames(Mnemiopsis_count_data, var = "gene")
rm(Mnemiopsis_count_data)
```

```{r}
# drop rows that rowsum =< 10
mncount$rowsum <- apply(mncount, 1, sum)
mncount <- subset(mncount, rowsum > 10)
mncount$rowsum <- NULL
dim(mncount)
```
#1) Build hierarchical trees based on the columns and for the rows (exclude rows that are "low" expression)
```{r}
# data preparation
# sampling for lambda calculation
sam <- NULL
for (c in 1:30){
  mnc <- sample(as.matrix(mncount), size = 100)
  sam <- append(sam, mnc)
}
opar <- par(mfrow = c(1, 2))
hist(sam); qqnorm(sam)
par(opar)
shapiro.test(sam)
```


```{r}
# define boxcox function
bxcx <- function(x, lmbd){
  if (abs(lmbd) < 0.05){
    return(log(x))
  }else{
    return((x^lmbd - 1)/lmbd)
  }
}

# for-loop to find lambda 
x <- sam
lam <- seq(-2, 2, 0.1)
t <- list()
v <- list()
cnt = 0
for (i in lam){
  cnt <- cnt + 1
  tt <- shapiro.test(bxcx(x,i))
  t[cnt] <- tt$statistic
  v[cnt] <- i
}
plot(v, t)
abline(v = 0.2)
```

```{r}
rm(sam)

# boxcox transfer 
mncount.bxcx <- bxcx(mncount, 0.2)
summary(mncount.bxcx)
```
```{r}
# test the distribution
sam2 <- sample(as.matrix(mncount.bxcx), size = 3000)
par(mfrow = c(1, 2))
hist(sam2); qqnorm(sam2)

shapiro.test(sam2)
```

```{r}
rm(sam2)
graphics.off()

# === scale transformation
mncount.t2 <- scale(mncount.bxcx, center = T, scale = T)
mncount.t2 <- as.data.frame(mncount.t2)
summary(mncount.t2)
```

```{r}
# built hierarchical tree
distmn2 <- dist(mncount.t2, method = "euclidean") 
hmn2 <- hclust(distmn2, method = "complete")
plot(hmn2, cex = 0.3, hang = -1)
```

# 2) Draw a heat map of the expression data
```{r}
rm(distmn2, hmn2)
graphics.off()
pheatmap(mncount.t2)
```

# 3) Use DESeq2 to analyze this data
# a. Which are the most significantly changing genes in this dataset?
```{r}
rm(mncount, mncount.bxcx, mncount.t2)
graphics.off()

# load data
countdata <- read.csv("Mnemiopsis_count_data.csv", header = TRUE, sep = ",")
metadata <- read.csv("Mnemiopsis_col_data.csv", header = TRUE, sep = ",")
head(countdata); head(metadata)
```
```{r}
# construct DESEQ dataset object
dds <- DESeqDataSetFromMatrix(countData = countdata,
                              colData = metadata,
                              design = ~condition,
                              tidy = TRUE)
dds
```
```{r}
# run DESEQ function
dds <- DESeq(dds)

# results table
res <- results(dds)
head(results(dds, tidy = TRUE))
```
```{r}
# sort summary list by p-value
res <- res[order(res$padj),]
head(res)
```
```{r}
# plot counts
par(mfrow = c(2,3))

plotCounts(dds, gene = "ML087114a", intgroup = "condition")
plotCounts(dds, gene = "ML20265a", intgroup = "condition")
plotCounts(dds, gene = "ML463533a", intgroup = "condition")
plotCounts(dds, gene = "ML085213a", intgroup = "condition")
plotCounts(dds, gene = "ML01433a", intgroup = "condition")
plotCounts(dds, gene = "ML01248a", intgroup = "condition")
```

# b. Which genes are most consistently highly expressed in these datasets that are the # "house-keeping" genes?
```{r}
res <- res[order(res$baseMean),]
tail(res) 
```
```{r}
# plot counts 
par(mfrow = c(2,3))

plotCounts(dds, gene = "ML04011a", intgroup = "condition")
plotCounts(dds, gene = "ML00017a", intgroup = "condition")
plotCounts(dds, gene = "ML020045a", intgroup = "condition")
plotCounts(dds, gene = "ML46651a", intgroup = "condition")
plotCounts(dds, gene = "ML26358a", intgroup = "condition")
plotCounts(dds, gene = "ML20395a", intgroup = "condition") 
```



