---
title: "mt2"
author: "Xiaoyan Wen"
date: "10/24/2020"
output: word_document
---

```{r}
library(dplyr)
library(tibble)
library(tidyr)
library(Hmisc)
```

#Data set 2: a) Mnemiopsis_col_data.csv b) Mnemiopsis_count_data.csv
```{r}
# read the data
m_count <- read.csv("Mnemiopsis_count_data.csv", header = T, stringsAsFactors=FALSE, row.names = 1)
str(m_count); head(m_count)
```

# 1) What are the top 5 genes with the highest average expression (across experiments) in the set ?
```{r}
# add one column as row_mean
m_count_sta <- mutate(m_count, row_mean = as.numeric(rowMeans(m_count)))  
# add row name
row.names(m_count_sta) <- row.names(m_count)
# select top 5 rows based on row mean value
m_count_topmean <- slice_max(m_count_sta, order_by = row_mean, n = 5)
# find out the row name for the top5
m_count_topmean_gene <- row.names(m_count_topmean)
m_count_topmean_gene
```

#2) Are the top 5 genes different if they are done on a per column basis ?

```{r}
m_count_c_topmean <- c()
for (i in c(1:8)){
  top5r <- slice_max(m_count_sta, order_by = m_count_sta[,i], n = 5)
  top5gene <- row.names(top5r)
  m_count_c_topmean <- rbind(m_count_c_topmean, top5gene)
}

m_count_c_topmean
```
# 3-1) Calculate mean and standard deviation of each column
```{r}
head(m_count_sta); str(m_count_sta); dim(m_count_sta)  # check the data

col_mean <- colMeans(m_count_sta)             # calculate column mean

col_sd <- c()                                # calculate column sd
for (i in c(1:ncol(m_count_sta))){
  csd <- sd(m_count_sta[,i])
  col_sd <- append(col_sd, csd)
}
col_sd

# combine the mean and variance into a new data frame "col_sta"
col_sta <- as.data.frame(rbind( mean = col_mean, sd = col_sd))  
colnames(col_sta) <- colnames(m_count_sta)     # add column name to the table
col_sta
```

# If the mean is different, then scale the columns so that they all have the same mean for the subsequent questions

```{r}
# calculate the scale_factor to make all the column means equal to row_mean
multiplier <- col_sta$row_mean[1]/col_sta[1,]

m_count_sta_b <- m_count_sta   # set a new data set copied from the original one
for (i in c(1: 9)){                    # loop to x multiplier for each column
  m_count_sta_b[,i] <- m_count_sta[,i]*multiplier[,i]
}

m_count_b <- m_count_sta_b[,c(1:8)]
summary(m_count_b)          # summary the new data set
```

# 3-2) Use correlations between columns to find the samples that are closely related. 
```{r}
result <- rcorr(as.matrix(m_count_b), type = "pearson") # find out correlations
c_r <- result$r                                   # check correlation r 

# select maximum value from the c_r upper triangle (not include diagonal)
top_c_r <- tail(sort(c_r[upper.tri(c_r)]), 1)

c_r <- as.matrix(c_r)
top_c_r_ind <- which(c_r == top_c_r, arr.ind = T) # find the corresponding index in the c_r
top_c_r_ind
```
# 4) Use correlations between rows to find the closest pairs (top 5)
```{r}

```


# 5) If you were forced to divide the genes in each column into high, medium and low count genes, how would you do this based on the data that you have ?
```{r}
par(mfrow = c(2,4))                   # plot histogram for each column
for (i in c(1:8)){
  hist(m_count_b[,i])
}
```

```{r}
x <- cut(m_count_b[,1], 3, include.lowest=TRUE, labels=c("Low", "Med", "High"))
table(x)                      # cutting according to raw_value is not working
```
```{r}
# before transfer the data to log format, drop rows having zeros
m_count_b2 <- m_count_b            
for (i in c(1:8)){
  m_count_b2 <- filter(m_count_b2, m_count_b2[,i] > 0)
}

dim(m_count_b2); summary(m_count_b2)
```

```{r}
# distribution of a log transformed dataset
m_count_b2_log <- log(m_count_b2)
par(mfrow = c(2,4))
for (i in c(1:8)){
  hist(m_count_b2_log[,i])
}
summary(m_count_b2_log)
```

```{r}
# cutting according to log_value and plot boxplot

par(mfrow = c(2,4))
x_tb <- c()

for (i in c(1:8)){
  x2 <- cut(m_count_b2_log[,i], 3, include.lowest=TRUE, labels=c("Low", "Med", "High"))
  x_tb <- rbind(x_tb,table(x2))
  boxplot(m_count_b2_log[,i] ~ x2)
}

x_tb

# would normalized the column mean, dropping the genes with zero, log the data and then group to "high", "medium" and "low".
```

# 6) make a list of the top 5 genes with most variability and top 5 genes with least variability (exclude genes that have low expression values)
```{r}
# calculate row_mean, row_sd and add as columns to the no_zero dataset "m_count_b2"
m_count_b2[,c(9:12)] <- NULL
m_count_b2 <- cbind(m_count_b2, row_mean = apply(m_count_b2, 1, mean), row_sd = apply(m_count_b2, 1, sd))
dim(m_count_b2); summary(m_count_b2)


top5_varia <- slice_max(m_count_b2, order_by = "row_sd", n = 5)
top5_varia[ , c(9:10)]

bot5_varia <- slice_min(m_count_b2, order_by = m_count_b2$row_sd, n = 5)
bot5_varia[ , c(9:10)]
```

# 7) Using the labels of columns provided, find the top variable genes between the two groups using a t-test
# list the 5 most up regulated and 5 most downregulated genes.
```{r}
# perform t test between the two group for each row, add columns for mean_ratio, p_value, FDR_value

# loop to perform t.test and extract p.value, estimate for each row;
t_esti <- c()
t_pval <- c()
ad_pval <- c()
for (i in c(1:dim(m_count_b2)[1])){
  t <- t.test(m_count_b2[i,c(1:4)], m_count_b2[i,c(5:8)])
  fdr <- p.adjust(t$p.value, method = "BH")
  t_esti <- rbind(t_esti, t$estimate)
  t_pval <- rbind(t_pval, t$p.value)
  ad_pval <- rbind(ad_pval, fdr)
}
m_count_b2[,c(11:13)] <- NULL 
m_count_b2 <- cbind(m_count_b2, mean_ratio = t_esti[,2]/t_esti[,1], t_pvalue = t_pval, FDR = ad_pval)
dim(m_count_b2); summary(m_count_b2)
```

```{r}
# select the top variable genes using criteria of FDR < 1%

topvar_from_t <- filter(m_count_b2, FDR < 0.01)
nrow(topvar_from_t)

# list the top5 up- and down- regulated genes 
top5_up <- slice_max(topvar_from_t, order_by = mean_ratio, n = 5)
top5_up
top5_down <- slice_min(topvar_from_t, order_by = mean_ratio, n = 5)
top5_down
```
























